### 2007-09-26 pkg/paper/article.Rnw
full set of one-possible-scenario conversions complete
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@7 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{article}
\usepackage{graphicx}
\begin{document}

@@

In this colorspace, Y corresponds roughly to how bright a color
appears, but the X and Z components have no simple intuitive 
meaning.  Furthermore, unit steps within this colorspace are
not perceptually equal, so generating a palette of colors cannot
proceed in a rational manner.  This leads us on to consideration 
to other colorspaces that are derived from CIE XYZ.

@@

The CIE LUV colorspace, technically referred to as CIE L*u*v*,
 is a transformation of CIE XYZ in a 
reasonably successful attempt to achieve perceptual uniformity.
Again, any color can be specified as a triplet of values,
with the advantage that a unit step in any direction corresponds to a 
fixed perceptual change in color.  However, the problem of 
intuitively specifying colors remains.

\subsection{Intuitive color choice:  CIE LCh}

The CIE LCh colorspace is a re-expression of CIE LUV using
polar coordinates for the U-V plane.  In this colorspace,
any color may be described by a triplet of values, a unit
step is perceptually uniform throughout the colorspace, 
and the three dimensions of the colorspace have intuitive 
interpretations:

\begin{description}
\item[L] corresponds to the luminance or
brightness of the color.  This is normalised to
the range 0 to 100.
\item[C] corresponds to the chroma (colorfulness, or purity) of the color.
Lowering the chroma produces a color that is closer to grey.
The range of this parameter depends on the luminance and hue.  For example,
the range of possible chromas is much smaller for 
very dark colors (low luminance);  in the extreme case of luminance 0,
which corresponds to black, chroma can only take the value 0.
\item[h] corresponds to the hue of the color.
This value can range from 0 to 360, with 0 corresponding to red, 
120 to green, 240 to blue, and so on.
\end{description}

@@

Only one problem remains:  we can conveniently specify a color
using CIE LCh, but in order to render that color we need a
specification that a rendering device understands.
For example, in order to display a color on screen,
we need to be able to convert a CIE LCh color into an RGB
color specification.

@@

The conversion from CIE LCh to RGB consists of a series
of steps.  We start with conversions between the various
CIE colorspaces.

\subsection{LCh to LUV}

@@

$$u* = C \times cos(h)$$

$$v* = C \times sin(h)$$

@@

This conversion is non-linear, reflecting the non-linear
nature of human color perception.

The conversion also depends on the concept of a \emph{white point}.
For computer monitors, the white point is what you get when you
combine red, green, and blue channels at full power.  This is not as
unambiguous as it sounds because it depends both on the SPDs 
of the red, green, and blue channels \emph{and} on the relative 
powerfulness of the maximum for each channel.  In effect, it is 
possible for two monitors to have very different white points, 
with one, for example, much ``bluer'' than another if the maximum
of the blue channel is set higher.\footnote{It may be easier
to think about a white point in terms of a printed image.  In that case,
the white point is an area of the image that perfectly reflects the
ambient lighting conditions.  In other words, the white point
corresponds to whatever illumination is being used to view the image.}

Typically, the white point for a computer monitor is set by the
manufacturer and the common setting used these days is something called 
\emph{D65}, which is a CIE specification of a colour that
corresponds roughly to the SPD of daylight.

@@

\subsection{D65}

The standard white point.

\subsection{Rec. 709}

The standard RGB chromaticities.

@@

The problem also remains:  given a CIE specification, how do we 
make a specific device emit that color.  Ergo transformations
between colorspaces.

>>>>

\documentclass{article}
\setlength{\parskip}{.1in}
\setlength{\parindent}{0in}
\usepackage{graphicx}
\begin{document}

@@

In this colorspace, $Y$ corresponds roughly to how bright a color
appears, but the $X$ and $Z$ components have no simple intuitive 
meaning.  Furthermore, unit steps within this colorspace are
not perceptually equal, so generating a palette of colors cannot
proceed in a rational manner.  This leads us on to consideration 
to other colorspaces that are derived from CIE XYZ.

@@

The CIE LUV colorspace
 is a transformation of CIE XYZ in a 
reasonably successful attempt to achieve perceptual uniformity.
Again, any color can be specified as a triplet of values,
with the advantage that a unit step in any direction corresponds to a 
fixed perceptual change in color.  However, the problem of 
intuitively specifying colors remains.

\subsection{Intuitive color choice:  CIE LCH}

The CIE LCH colorspace is a re-expression of CIE LUV using
polar coordinates for the U-V plane.  In this colorspace,
any color may be described by a triplet of values, a unit
step is perceptually uniform throughout the colorspace, 
and the three dimensions of the colorspace have intuitive 
interpretations:

\begin{description}
\item[$L$] corresponds to the luminance or
brightness of the color.  This is normalised to
the range 0 to 100.
\item[$C$] corresponds to the chroma (colorfulness, or purity) of the color.
Lowering the chroma produces a color that is closer to grey.
The range of this parameter depends on the luminance and hue.  For example,
the range of possible chromas is much smaller for 
very dark colors (low luminance);  in the extreme case of luminance 0,
which corresponds to black, chroma can only take the value 0.
\item[$H$] corresponds to the hue of the color.
This value can range from 0 to 360, with 0 corresponding to red, 
120 to green, 240 to blue, and so on.
\end{description}

@@

Only one problem remains:  we can conveniently specify a color
using CIE LCH, but in order to render that color we need a
specification that a rendering device understands.
For example, in order to display a color on screen,
we need to be able to convert a CIE LCH color into an RGB
color specification.

@@

The conversion from CIE LCH to RGB consists of a series
of steps.  We start with conversions between the various
CIE colorspaces.

\subsection{LCH to LUV}

@@

$$U = C \cos(H)$$

$$V = C \sin(H)$$

@@

This conversion is non-linear, reflecting the non-linear
nature of human color perception.  It is also performed in 
several stages, reflecting the historical development of the
CIE LUV colorspace, which builds upon a couple of other
earlier transformations of the CIE XYZ colorspace.

The conversion also depends on the concept of a \emph{white point}.
For now, we can just think of the white point as representing the
viewing conditions under which the colour is to be shown.
\footnote{It is relatively straightforward 
to think about a white point in terms of a printed image.  In that case,
the white point is an area of the image that perfectly reflects the
ambient lighting conditions.  In other words, the white point
corresponds to whatever illumination is being used to view the image.}
The white point is important because the perception of colors
is dependent on the viewing conditions.  For example, an image
will look different when viewed under a pink light instead
of a white light.  More subtly, 
an image will look different when viewed outside in normal sunlight compared
to inside with fluorescent lighting.

The white point is expressed as CIE XYZ tristimulus values.
There are a number of established white points used in practice,
with the most popular being a representation of normal daylight,
called $D65$:

% Source:  Ross's code, confirmed by published article

$$\begin{array}{c}X_r = 95.047\\
                  Y_r = 100.000\\
                  Z_r = 108.883
  \end{array}$$

\textit{Put more of these in an appendix?}

% Source:  http://brucelindbloom.com/

The value of $Y$ is relatively straightforward to calculate from the
value of $L$:

$$Y = \left\{\begin{array}{l l}((L + 16)/116)^3 & \quad L > 7.999592 \\
                               L/903.3 & \quad L \leq 7.999592 \\
             \end{array} 
      \right.$$

Before we can calculate $X$ and $Z$, we need to calculate several 
intermediate values:

$$\begin{array}{l}u_r = \frac{4X_r}{X_r + 15Y_r + 3Z_r}\\
                  v_r = \frac{9Y_r}{X_r + 15Y_r + 3Z_r}\\
  \end{array}$$

$$\begin{array}{l}a = \frac{1}{3}\left(\frac{52L}{U + 13Lu_r} - 1\right)\\
                  b = -5Y\\
                  c = -\frac{1}{3}\\
                  d = Y\left(\frac{39L}{V + 13Lv_r} - 5\right)\\
  \end{array}$$

\textit{Check me: should those be straight 
U and V, or ``normalised'' versions?}

The values of $X$ and $Y$ are now calculated from these values:

$$X = \frac{d - b}{a - c}$$

$$Z = Xa + b$$

@@

\textit{Fixme:  need to deinfe ``primaries'' earlier.}

One such standard is the ITU-R BT.709-2 Recommendation (for
high-definition television), which specifies CIE XYZ values
for the $R$, $G$, and $B$ primaries of a display device.

\textit{Fixme:  need to deinfe ``channels'' earlier.}

The second point is that we again need a white point in
order to perform the conversion.
For computer monitors, the white point is what you get when you
combine red, green, and blue channels at full power.  This is not as
unambiguous as it sounds because it depends both on the SPDs 
of the red, green, and blue channels \emph{and} on the relative 
powerfulness of the maximum for each channel.  In effect, it is 
possible for two monitors to have very different white points, 
with one, for example, much ``bluer'' than another if the maximum
of the blue channel is set higher.

Typically, the white point for a computer monitor is set by the
manufacturer and the common setting used these days is 
D65 (as described in the previous section).

Using the 709-2 Recommendation primaries and a D65 white point,
the conversions are:

% source:  Poynton's Color FAQ

\textit{Check me again:  are we dealing with tristimulus values
or chromaticities?  Ross's code says chromaticities, but Bruce
Lindbloom's site seems to allow working in tristimulus values.}

$$R = 3.240479X - 1.537150Y - 0.498535Z$$

$$G = -0.969256X + 1.875992Y + 0.041556Z$$

$$B = 0.055648X - 0.204043Y + 1.057311Z$$

\textit{Discuss calculation of transforms from RGB primaries that
are not Rec. 709?}

\textit{Need to talk about clipping to the RGB gamut}

\section{Gamma correction}

@@

The sRGB colorspace consists of colour specifications as 
RGB tristimulus values, \emph{plus} fixed CIE specifications of
the $R$, $G$, and $B$ primaries, \emph{plus} a fixed white point, 
\emph{plus} a fixed gamma.
This is in effect a device-independent colorspace because it
dictates the characteristics of a display device (and the viewing
conditions).
